\chapter*{摘\quad 要}
\addcontentsline{toc}{chapter}{摘要}

胸部X光图像肺炎分类是医学图像模式识别中的典型二分类问题，也常被用于迁移学习课程实验。公开数据集上的较高分数并不自动表示模型具有稳定的跨人群或跨机构能力：训练集与测试集可能共享患者，验证集可能过小，外部数据的标签语义也可能不同。若不先处理这些问题，继续增加模型或调节阈值会放大不可靠结论。

本课程设计建立了覆盖数据完整性审计、患者级重划分、多模型训练、外部评价、概率校准、错误分析和网页演示的可复现实验流程。项目首先审计Kermany儿童胸片数据的5856张图像。字节级哈希未发现精确图像跨集合重复，但依据文件名解析的肺炎患者编号有170个同时存在于官方训练集与测试集。为保证评价独立性，项目汇总原有目录后按患者分组，重新划分为4107张训练图像、579张验证图像和1170张锁定内部测试图像；三个集合的患者标识完全隔离。RSNA镜像可用子集包含1707张图像，测试集为442张；NIH弱标签子集包含130张图像，仅用于敏感性分析。

严格基线比较DenseNet121、ConvNeXt-Tiny与ViT-B/16，每种模型运行三个随机种子。三模型集成在患者级内部测试集上的准确率为0.9778--0.9821、ROC-AUC为0.9945--0.9968，95\%患者级bootstrap区间高度重叠。直接迁移至RSNA后，准确率下降到0.6674--0.6765，平衡准确率下降到0.6591--0.6695，说明数据来源差异的影响远大于三种结构之间的差异。仅使用亮度、灰度分布、边缘和中心区域统计量的来源分类器，其五折组间验证AUC达到0.9551，表明两个数据来源存在明显、可测量的图像风格差异。

低成本稳健训练在DenseNet121上加入轻度强度扰动和0.1标签平滑。三个随机种子的RSNA平衡准确率均值由严格基线提高到0.7116，内部平衡准确率保持在0.9742；但RSNA AUC降至0.7826，说明该方案改善了固定阈值行为和概率分布，却没有全面改善排序能力。项目据此把结果表述为部分改善，并继续比较患者级混合训练与域均衡采样。所有阈值与温度参数只允许从验证数据学习，测试集阈值扫描仅用于事后说明误报和漏检的权衡。

实验表明，课程设计最重要的提升并非增加网络数量，而是把数据独立性、外部标签差异、随机波动和概率可靠性纳入同一证据链。本项目只研究公开胸片的图像级二分类，不构成临床诊断、病灶定位或治疗依据。

\noindent\textbf{关键词：}胸部X光；肺炎分类；数据泄漏；患者级划分；外部验证；概率校准；域偏移

\chapter*{Abstract}
\addcontentsline{toc}{chapter}{Abstract}

Chest radiograph pneumonia classification is a representative binary pattern-recognition task, but high performance on a public split does not by itself imply reliable generalization. This project develops an auditable pipeline covering data-integrity checks, patient-grouped splitting, multi-model training, external evaluation, calibration, error analysis, and a local web demonstration. An audit of 5,856 Kermany images found no exact cross-split duplicate files but identified 170 pneumonia subject identifiers shared by the official training and test folders. A new patient-grouped split was therefore frozen with 4,107 training, 579 validation, and 1,170 locked internal-test images.

DenseNet121, ConvNeXt-Tiny, and ViT-B/16 were trained with three random seeds. Their ensemble accuracies on the patient-independent internal test set ranged from 0.9778 to 0.9821, whereas accuracy on the 442-image RSNA external test set fell to 0.6674--0.6765. A source classifier based only on simple image statistics achieved a grouped cross-validation AUC of 0.9551, confirming measurable acquisition and style differences. A low-cost robust DenseNet training recipe improved mean external balanced accuracy to 0.7116 while preserving internal performance, but reduced external ROC-AUC; it is therefore interpreted as a partial operating-point improvement rather than a universal gain. The work demonstrates why split integrity, uncertainty, and external label semantics must be examined before model rankings are accepted. The system is intended solely for coursework and research training and is not a clinical diagnostic tool.

\noindent\textbf{Keywords:} chest radiograph; pneumonia classification; data leakage; patient-level split; external validation; calibration; domain shift
